


%----- Abstract
\input{frontbackmatter/abstract}


\renewcommand{\cftchapterfont}{\normalsize\normalfont}   
\renewcommand{\cftpartfont}{\normalsize\normalfont}   
\renewcommand{\cftsectionfont}{\normalsize\normalfont} 
\renewcommand{\cftsubsectionfont}{\normalsize\normalfont} 
\renewcommand*{\cftpartname}{Part\space}
\newpage
\setcounter{tocdepth}{3} 
%\cftsetindents{chapter}{1.5em}{3.0em}
%\cftsetindents{section}{3em}{3.0em}
%\cftsetindents{subsection}{4.5em}{3.0em}
\tableofcontents*
\cleardoublepage
\listoffigures
\cleardoublepage
\listoftables
\cleardoublepage
% \listofalgorithms

% \listoflistings

%\printnomenclature % Notation stuff in main tex file

\clearpage
\chapter{Acknowledgements}

The University of Washington Department of Anthropology, with whom I have been fortunate to be associated in one form or another since 1985, provided financial support, lab space, computer time (back in the days when you paid by the hour!), and other resources needed to conduct this research.  Most of all, the Department tolerated and supported the unusual and circuitous path my graduate studies and research have taken.  Thank you for your support.  I was introduced to the Department by Stevan Harrell through his participation in the UW Honors Program.  Faculty with whom I had significant interaction and who influenced me include Jim Green, David Spain, Stevan Harrell, Miriam Kahn, Laura Newell, Steven Goodreau, Gerald Eck, Eric Alden Smith, Julie K. Stein, Donald K. Grayson and Robert Wenke.  Thank you all for a superb education.  I also wish to thank Peter D. Ward, whose graduate paleobiology course in the spring quarter of 1987 substantially increased my understanding of evolutionary biology and broadened my research horizons.  

\vskip 0.3cm

I thank everyone who has served on both incarnations of my dissertation committee, including Robert C. Dunnell, Donald K. Grayson, Thomas Stoebe, Carl Bergstrom, Carl Lipo, Ben Marwick, and James K. Feathers.  In particular, Jim Feathers and Ben Marwick provided strong encouragement at a key moment, and my gratitude for that push is eternal.  Thank you both.

\vskip 0.3cm

Carl Lipo has been a friend, colleague, and much more since 1988.  Our collaboration is strongly represented in these pages, and I look forward to many more years of thinking about these issues and collaborative work together.  

\vskip 0.3cm

Over the years, my colleagues from the Department have become friends, many lifelong and beyond.  I want to specifically thank Sarah Sterling, Fran Hamilton, Michael Pfeffer, Betsy Scharf, Steve Cole, and Chris Pierce.  Most especially, Kim Kornbacher left us before I could finish this and share the joy with her, but she lives forever in my heart.  Kris Wilhelmsen---please know that you and Kim are forever family to me!

\vskip 0.3cm

A very special thanks goes to Lara Braithwaite, life partner and friend, for sustaining me through the latter stages of this project with great love, encouragement, and steadfast support.  Thank you for our life together! 
\vskip 0.3cm

Finally, I thank Robert Chester Dunnell for the gift of his teaching, his thought, his insatiable curiosity, and his discipline.  Dr. Dunnell was the very model of a scientist and scholar, and his influence can be seen on every page.  From the moment I walked into his Archy 497 classroom in the autumn of 1986, he has been an inspiration, and set a standard to which I continue to try to attain.  



\input{frontbackmatter/dedication}

% \newpage 

% \input{frontbackmatter/quote}



\newpage
%\input{content/preface}